Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-18 paper_jasa.tex
figures 10 and 11 - changed layers

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate, 
% If there is a signal in the data plot, visual $p$-value has a tendency to take a smaller value (small enough to reject $H_0$) and if not it takes a larger value (large enough to not reject $H_0$) 
% 
 giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment 2. This guides the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values for which researchers get confused whether to reject or 'not reject' $H_0$. For visual test this is not common.

@@

We also see the same pattern from the loess smoother fitted through the points in Figure \ref{fig:pval_definition}. The sigmoid shaped loess curves indicate a clear separation of visual $p$-values. For experiment 1, most of the visual $p$-values are above 0.4 when conventional $p$-values are above 0.2. This is more evident for experiment 2 where most of the visual $p$-values are above 0.8. For experiment 3, we see all the visual $p$ values are almost zero no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject and visual test does the job nicely.

@@

One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the plot with the lowest $p$-value in the lineup. (Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value.) In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. But even when they didn't pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=5, n=300, \sigma=12, rep=3$, people tended to pick the plot with the second smallest $p$-value.  Same experiment, when $\beta=0, n=100, \sigma=5, rep=1$ people overwhelming chose a plot with much larger $p$-value. For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine what the observers were looking at in making their choices \citep{zhao:2012}.
(Lineups chosen for the eye-tracker experiment were in experiment 1, (1) $n=100, \beta=3, \sigma=5, rep=1$ (2) $n=100, \beta=5, \sigma=5, rep=1$ (3) $n=100, \beta=10, \sigma=12, rep=1$ and (4) $n=300, \beta=5, \sigma=5, rep=1$, and in experiment 2, (5) $n=100, \beta=3.5, \sigma=12, rep=1$ (6) $n=100, \beta=4.5, \sigma=12, rep=2$ (7) $n=300, \beta=1, \sigma=5, rep=1$ and (8) $n=300, \beta=1, \sigma=5, rep=2$.)

>>>>

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate, 
% If there is a signal in the data plot, visual $p$-value has a tendency to take a smaller value (small enough to reject $H_0$) and if not it takes a larger value (large enough to not reject $H_0$) 
% 
 giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment 2. This guides the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values making the decision whether to reject or not harder.
 % for which researchers get confused whether to reject or 'not reject' $H_0$. 
 For visual tests this is not common.

@@

%We also see the same pattern from the loess smoother fitted through the points in Figure \ref{fig:pval_definition}. 
%The sigmoid shaped loess curves indicate a clear separation of visual $p$-values. For experiment 1, most of the visual $p$-values are above 0.4 when conventional $p$-values are above 0.2. This is more evident for experiment 2 where most of the visual $p$-values are above 0.8.

For experiment 3, we see that  the visual $p$ values are very small no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject, whereas the visual test performs well.
% and the visual test does the job nicely.

@@

\noindent One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

%Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.)

Red indicates the plot with the lowest $p$-value in the lineup. Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value. In both experiments  people tended to select the plot with the lowest $p$-value with an increase in effect size.
 %This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. 
 The results are clearer for experiment 2, that used a continuous covariate. But even when subjects did not pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=0, n=100, \sigma=5, rep=1$ people overwhelmingly chose a plot with much larger $p$-value, similarly, for 
parameters $\beta=5, n=300, \sigma=12, rep=3$, people tended to pick the plot with the second smallest $p$-value.
For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine what the observers were looking at in making their choices \citep{zhao:2012}.
%(Lineups chosen for the eye-tracker experiment were in experiment 1, (1) $n=100, \beta=3, \sigma=5, rep=1$ (2) $n=100, \beta=5, \sigma=5, rep=1$ (3) $n=100, \beta=10, \sigma=12, rep=1$ and (4) $n=300, \beta=5, \sigma=5, rep=1$, and in experiment 2, (5) $n=100, \beta=3.5, \sigma=12, rep=1$ (6) $n=100, \beta=4.5, \sigma=12, rep=2$ (7) $n=300, \beta=1, \sigma=5, rep=1$ and (8) $n=300, \beta=1, \sigma=5, rep=2$.)
